Explore expression of marker genes for tumor cell states

### 2025-01-10 analyses/cell-type-ewings/exploratory_analysis/06-tumor-cell-state-assignment.Rmd
update umaps and heatmaps

    p1 <- cluster_classification_df |>
      dplyr::filter(library_id == id) |>
      ggplot(aes(x = UMAP1, y = UMAP2, color = singler_lumped)) +
      geom_point(alpha = 0.5, size = 0.1) +
      theme(
        aspect.ratio = 1
      ) +
      scale_color_brewer(palette = "Dark2") +
      labs(title = id) +
      guides(color = guide_legend(override.aes = list(alpha = 1, size = 1.5)))

    p2 <- cluster_classification_df |>
      dplyr::filter(library_id == id) |>
      ggplot(aes(x = UMAP1, y = UMAP2, color = as.factor(cluster))) +
      geom_point(alpha = 0.5, size = 0.1) +
      theme(
        aspect.ratio = 1
      ) +
      #scale_color_brewer(palette = "Dark2") +
      labs(title = id, color = "cluster") +
      guides(color = guide_legend(override.aes = list(alpha = 1, size = 1.5)))

@@

```{r, fig.height=7, fig.width=10}
# pull out columns that have sum of marker gene expression 
marker_gene_cols <- colnames(cluster_classification_df)[stringr::str_detect(colnames(cluster_classification_df), "_sum$")]

# look at the expression of all gene sets across both libraries 
marker_gene_cols |>
  purrr::map(\(geneset){
    cluster_classification_df |>
      ggplot(aes(x = UMAP1, y = UMAP2, color = !!sym(geneset))) +
      geom_point(alpha = 0.5, size = 0.1) +
      facet_wrap(vars(library_id)) +
      scale_color_viridis_c() +
      theme(
        aspect.ratio = 1
      )
  }) |> 
  patchwork::wrap_plots(ncol = 2)
```

@@

- It doesn't look like very many cells express proliferative markers, so perhaps that gene set will not be helpful for these libraries? 
- There are some very obvious `EWS-FLI1` low groups of cells. 
- The `EWS-high` and `aynaud-EWS-high` show clusters that have high expression of `EWS-FLI1` targets, but the `aynaud-EWS-high` appears to have some more separation across clusters. 
That tells me that gene signature might be more helpful in classifying cells across the `EWS-FLI1` continuum.

@@

Now we will look at the marker gene expression across all clusters for each sample, first using a density plot and then a heatmap.

@@

@@

    full_celltype_heatmap(filtered_df, marker_gene_cols, "sub_cluster", plot_name = id)

@@

@@

Between these two plots I think we can try and group cells into EWS low, EWS middle, and EWS high cells. 
I don't think the proliferative markers are very informative, so I don't think we can use that to categorize any cells.

@@

`SCPCL000824` : 
`EWS-FLI1` low - 1
`EWS-FLI1` middle - 6 
`EWS-FLI1` high - 2, 3, 4, 5, 7

```{r}
# categorize tumor cells based on EWS expression 
tumor_cells_df <- tumor_cells_df |> 
  dplyr::mutate(
    cell_state = dplyr::case_when(
      library_id == "SCPCL000822" & sub_cluster == 2 ~ "EWS-low",
      library_id == "SCPCL000822" & sub_cluster %in% c(1, 4) ~ "EWS-mid",
      library_id == "SCPCL000822" & sub_cluster == 3 ~ "EWS-high",
      library_id == "SCPCL000824" & sub_cluster == 1 ~ "EWS-low",
      library_id == "SCPCL000824" & sub_cluster %in% c(6, 8) ~ "EWS-mid",
      library_id == "SCPCL000824" & sub_cluster %in% c(2, 3, 4, 5, 7) ~ "EWS-high",
    )
  )
```

Let's re-plot the heatmap but label by tumor cell state rather than cluster.

@@

    full_celltype_heatmap(filtered_df, marker_gene_cols, "cell_state", plot_name = id)

>>>>

    p1 <- umap_df |>
      dplyr::filter(library_id == id) |>
      ggplot(aes(x = UMAP1, y = UMAP2, color = singler_lumped)) +
      geom_point(alpha = 0.5, size = 0.1) +
      theme(
        aspect.ratio = 1
      ) +
      scale_color_brewer(palette = "Dark2") +
      labs(title = id) +
      guides(color = guide_legend(override.aes = list(alpha = 1, size = 1.5)))

    p2 <- umap_df |>
      dplyr::filter(library_id == id) |>
      ggplot(aes(x = UMAP1, y = UMAP2, color = as.factor(cluster))) +
      geom_point(alpha = 0.5, size = 0.1) +
      theme(
        aspect.ratio = 1
      ) +
      #scale_color_brewer(palette = "Dark2") +
      labs(title = id, color = "cluster") +
      guides(color = guide_legend(override.aes = list(alpha = 1, size = 1.5)))

@@

```{r}
# pull out columns that have sum of marker gene expression 
marker_gene_cols <- colnames(cluster_classification_df)[stringr::str_detect(colnames(cluster_classification_df), "_sum$")]

# look at the expression of all gene sets across both libraries 
marker_gene_cols |>
  purrr::map(\(geneset){
    cluster_classification_df |>
      ggplot(aes(x = UMAP1, y = UMAP2, color = !!sym(geneset))) +
      geom_point(alpha = 0.5, size = 0.1) +
      facet_wrap(vars(library_id)) +
      scale_color_viridis_c() +
      theme(
        aspect.ratio = 1
      )
  })
```

@@

- It doesn't look like very many cells express proliferative markers in `SCPCL000822`, but cluster 2 has high expression in `SCPCL000824`. 
It looks like that might coincide with high `EWS-FLI` seen in that same cluster, but we have to dive deeper to know for sure. 
- There are some very obvious `EWS-FLI1` low groups of cells. 
- The `EWS-high` and `aynaud-EWS-high` show clusters that have high expression of `EWS-FLI1` targets, but the `aynaud-EWS-high` appears to have some more separation across clusters. 
That tells me that gene signature might be more helpful in classifying cells across the `EWS-FLI1` continuum.

@@

Now we will look at the marker gene expression across all clusters for each sample, first using a density plot and then heatmaps.

@@

In all of the heatmaps below, the `annotation` column is the `sub_cluster`.

@@

    full_celltype_heatmap(filtered_df, marker_gene_cols, "sub_cluster")

@@

Let's make the same heatmap, but remove the column clustering.

    filtered_df <- tumor_cells_df |> 
      dplyr::filter(library_id == id) |> 
      dplyr::arrange(sub_cluster)

    full_celltype_heatmap(filtered_df, marker_gene_cols, "sub_cluster", cluster_columns = FALSE)

@@

Between these two plots I think we can try and group cells into EWS low, EWS middle, and EWS high cells.

The last thing we'll do is look at any cells that could be marked as "proliferative". 
`SCPCL000822` doesn't appear to have very much expression of the proliferative markers, but `SCPCL000824` does seem to have one cluster, cluster 2, that has expression of the proliferative cells. 
Perhaps we should label those cells separately as `EWS-high-proliferative`.

@@

`SCPCL000824` : 
`EWS-FLI1` low - 1
`EWS-FLI1` middle - 6 
`EWS-FLI1` high - 3, 4, 5, 7
`EWS-FLI1` high proliferative - 2

```{r}
# categorize tumor cells based on EWS expression 
tumor_cells_df <- tumor_cells_df |> 
  dplyr::mutate(
    cell_state = dplyr::case_when(
      library_id == "SCPCL000822" & sub_cluster == 2 ~ "EWS-low",
      library_id == "SCPCL000822" & sub_cluster %in% c(1, 4) ~ "EWS-mid",
      library_id == "SCPCL000822" & sub_cluster == 3 ~ "EWS-high",
      library_id == "SCPCL000824" & sub_cluster == 1 ~ "EWS-low",
      library_id == "SCPCL000824" & sub_cluster == 6 ~ "EWS-mid",
      library_id == "SCPCL000824" & sub_cluster == 2 ~ "EWS-high-proliferative",
      library_id == "SCPCL000824" & sub_cluster %in% c(3, 4, 5, 7) ~ "EWS-high",
    )
  )
```

Let's re-plot the heatmap but label by tumor cell state rather than cluster. 
In this heatmap the annotation column is `cell_state`.

@@

    full_celltype_heatmap(filtered_df, marker_gene_cols, "cell_state")
